Linking hypothesis and number of response options modulate inferred scalar implicature rate

### 2018-10-31 manuscript.tex
updated gd

The linking hypothesis under discussion assumes that speaker
probabilities of utterance given meaning are invariant across a) our
four different experimental conditions, b) across participants, and c)
within participants (that is, participants are not capable of updating
their \(S_1\) distribution in a local discourse context). We note that
the assumption (b) may conceivably be relaxed by allowing one or more of
the parameters in the model -- including the prior probability over
world states \(P_w\), the cost function on utterances \(C\), or the
rationality parameter \(\alpha\) -- to vary across participants. We also
note that assumption (c) in particular is in tension with a growing body
of empirical evidence that semantic and pragmatic interpretation is
modulated by rapid adaptation to the linguistic and social features of
one's interlocutors (Fine, Jaeger, Farmer, \& Qian, 2013; Kleinschmidt
\& Jaeger, 2015).

However, if we should like to keep the above assumptions in place, then
we must look elsewhere to explain the observed variation in our
experimental data. In particular, this linking hypothesis, coupled with
our assumptions, commits us to explaining variation in the data in terms
of the threshold parameters of our responder model \(R\). Consider first
the variation in response across different experimental conditions on a
given trial, e.g.~evaluation of a guess of \enquote{cat and dog} when
the card contains both a cat and a dog. The variation in the proportion
of responses of \enquote{right} on this trial between the binary,
ternary, quaternary, and quinary conditions indicates that the threshold
value for \enquote{right} responses must vary across conditions; that
is, we predict that the \(\theta\) of the binary condition will differ
from, e.g., the \(\theta_1\) of the ternary condition as well as the
\(\theta_1\) of the quaternary condition. We also observed variation in
response on this trial within a single condition (for example, a
sizeable minority of participants responded \enquote{wrong} to this
trial in the binary condition). Thus, this linking hypothesis is
committed to the notion that threshold values may vary across
participants, such that a speaker probability of utterance \(S_1\)(u
\textbar{} s) can fall below \(\theta\) for some subset of participants
while \(S_1\)(u \textbar{} s) itself remains constant across
participants.

Lastly, it is conceivable that for two utterances of the same
conditional probability and in the same experimental condition, a
participant in our experiment provided a judgment of, e.g.
\enquote{right} to one utterance but \enquote{wrong} to the other. That
is, it is possible that there was within-subject variation in this
experiment. One way to represent such variation would be to posit that
the parameterization of threshold values proceeds stochastically and
that threshold values are recalibrated for every individual sentence
verification task. Rather than representing a threshold as a discrete
value N between 0 and 1, we can represent that threshold as a
distribution over possible threshold values -- with mass centered around
N. Whenever an individual encounters a sentence verification task, such
as a single trial of our truth value judgment task experiment, the
threshold value is recalibrated by sampling from this distribution. If
we allow values of \(\theta\) to vary as a result of this schotastic
process, for the possibility that \(S_1\)(u \textbar{} s) sometimes
falls below \(\theta\) (and is otherwise above \(\theta\)) for a given
participant.

One outstanding empirical problem is the pattern of response we observed
for \enquote{cat and dog} on trials where there was only a cat on the
card. Because this utterance is strictly false in this world state, it
is surprising -- on both the traditional view as well as on the account
developed here -- that participants assigned this utterance ratings
above \enquote{wrong} with any systematicity. However, this is exactly
what we observed, particulary in the quaternary and quinary conditions
of the experiment, where a sizeable minority of participants considered
this utterance \enquote{kinda right}. As Figure \ref{fig:speakerprobs}
demonstrates, the conditional speaker probability of this utterance in
this world state is 0; thus, there is no conceivable threshold value
that would allow this utterance to ever be rated above \enquote{wrong}
(on the reasonable assumption that the thresholds in our responder model
\(R\) should be nonzero). Any linking hypothesis will have to engage
with this data point, and we leave to future work an analysis which
captures participants' behavior in this condition.

For the time being, however, we present the above analysis as a proof of
concept for the following idea: by relaxing the assumptions of the
traditional view of scalar implicature (namely, that scalar implicatures
either are or are not calculated, and that behavior on sentence
verification tasks directly reflects this binary interpretation
process), we can propose quantitative models of the variation in
behavior we observe in experimental settings. We note that the linking
analysis proposed here is just one in the space of possible analyses
when traditional assumptions about scalar implicature are relaxed. For
example, one might reject this threshold-based analysis in favor of one
whereby responses are the outcomes of sampling on the (pragmatic speaker
or pragmatic listener) probability distributions provided by an RSA
model. We must leave this investigation to future work, but for now we
emphasize that this kind of quantitative, data-driven and systematic
model criticism is made available to researchers in experimental
pragmatics by revising core assumptions about the nature of scalar
implicature. Though we no longer have a crisp notion of scalar
implicature as something that is or is not \enquote{calculated} in
interpretation, we have new flexibility to explicitly discuss
categorical behavior in experimental settings.

@@

>>>>

The linking hypothesis under discussion assumes that speaker
probabilities of utterance given meaning are invariant across a) our
four different experimental conditions, b) across participants, and c)
within participants (that is, participants do not update
their \(S_1\) distribution in a local discourse context). We note that
the assumption (b) may conceivably be relaxed by allowing one or more of
the parameters in the model -- including the prior probability over
world states \(P_w\), the cost function on utterances \(C\), or the
rationality parameter \(\alpha\) -- to vary across participants. We also
note that assumption (c) in particular is in tension with a growing body
of empirical evidence that semantic and pragmatic interpretation is
modulated by rapid adaptation to the linguistic and social features of
one's interlocutors (Fine, Jaeger, Farmer, \& Qian, 2013; Kleinschmidt
\& Jaeger, 2015; \change{Yildirim, Degen, Tanenhaus, \& Jaeger, 2016)}.

However, if we should like to keep the above \change{simplifying} assumptions in place, then
this linking hypothesis commits us to explaining variation in the data in terms
of the threshold parameters of our responder model \(R\). Consider first
the variation in response across different experimental conditions on a
given trial, e.g.~evaluation of a guess of \enquote{cat and dog} when
the card contains both a cat and a dog. The variation in the proportion
of responses of \enquote{right} on this trial between the binary,
ternary, quaternary, and quinary conditions indicates that the threshold
value for \enquote{right} responses must vary across conditions; that
is, we predict that the \(\theta\) of the binary condition will differ
from, e.g., the \(\theta_1\) of the ternary condition as well as the
\(\theta_1\) of the quaternary condition. We also observed variation in
response on this trial within a single condition (for example, a
sizeable minority of participants responded \enquote{wrong} to this
trial in the binary condition). Thus, this linking hypothesis is
committed to the notion that threshold values may vary across
participants, such that a speaker probability of utterance \(S_1\)(u
\textbar{} s) can fall below \(\theta\) for some subset of participants
while \(S_1\)(u \textbar{} s) itself remains constant across
participants.

Lastly, for two utterances of the same
conditional probability and in the same experimental condition, 
participants in our experiment sometimes provided a judgment of
\enquote{right} to one utterance but \enquote{wrong} to the other. That
is, there was within-subject variation in this
experiment. One way to represent such variation would be to posit that
the parameterization of threshold values proceeds stochastically and
that threshold values are recalibrated for every individual sentence
verification task. Rather than representing a threshold as a discrete
value N between 0 and 1, we can represent that threshold as a
distribution over possible threshold values -- with mass centered around
N. Whenever an individual encounters  a single trial of our truth value judgment task experiment, a
threshold value is sampled from this distribution. \change{By  allowing values of \(\theta\) to vary stochastically in this way, we can capture that \(S_1\)(u \textbar{} s) can fall both above and below \(\theta\) for a given
participant.}

\change{The model in its present form already captures an interesting asymmetry in inferred implicature rates between exhaustive and scalar trials of the experiment: note specifically (c.f. Figure 8) that inferred implicature rates are greater in the binary and ternary conditions for scalar trials over exhaustive trials. This is expected given the model's inferred speaker probabilities: the speaker probability of producing ``There is a cat on the card'' in the context of there being a cat and dog on the card (an exhaustive implicature-inducing trial) is greater than the speaker probability of producing ``There is a cat or a dog on the card'' in that same context (a scalar implicature-inducing trial). Assuming noisy  $\theta$ values centered around N, participants are expected to respond `Right' more frequently on exhaustive than on scalar trials, which is precisely what is observed. Recall that these probabilities were derived via the simplifying assumption of uniform cost on utterances; in fact, adding cost to relatively complex disjunctive sentences over simple declarative sentences only predicts a more pronounced asymmetry in the experimentally-observed direction.}

%\change{While we have offered only a preliminary qualitative sketch of the model's behavior here, future work should systematically analyze the quantitative fit between model predictions and the data. }

%\change{Less readily explained is why this asymmetry disappears with the introduction of more response options, i.e. in the quaternary and quinary conditions. If anything, the opposite numerical trend is observed in these conditions: that is, inferred implicature rates are slightly greater in exhaustive trials over scalar trials. Future work should investigate the interaction between trial type (exhaustive vs. scalar) and number of response options.}

One  empirical problem is the pattern of responses we observed
for \enquote{cat and dog} on trials where there was only a cat on the
card. Because this utterance is strictly false in this world state, it
is surprising -- on both the traditional view as well as on the account
developed here -- that participants assigned this utterance ratings
above \enquote{wrong} with any systematicity. However, this is 
what we observed, particulary in the quaternary and quinary conditions
of the experiment, where a sizeable minority of participants considered
this utterance \enquote{kinda right}. As Figure \ref{fig:speakerprobs}
demonstrates, the conditional speaker probability of this utterance in
this world state is 0; thus, there is no conceivable threshold value
that would allow this utterance to ever be rated above \enquote{wrong}
(on the reasonable assumption that the thresholds in our responder model
\(R\) should be nonzero). Any linking hypothesis will have to engage
with this data point, and we leave to future work an analysis which
captures participants' behavior in this condition.

For the time being, however, we present the above analysis as a proof of
concept for the following idea: by relaxing the assumptions of the
traditional view of scalar implicature---namely, that scalar implicatures
either are or are not calculated, and that behavior on sentence
verification tasks directly reflects this binary interpretation
process---we can propose quantitative models of the variation in
behavior that is observed in experimental settings. We note that the linking
hypothesis proposed here is just one in the space of possible hypotheses. For
example, one might reject this threshold-based analysis in favor of one
whereby responses are the outcomes of sampling on the (pragmatic speaker
or pragmatic listener) probability distributions provided by an RSA
model. We \change{leave this systematic, quantitative} investigation to future work. For now we
emphasize that \change{explicit computational modeling of behavioral responses is a tool that is  available to researchers in experimental
pragmatics. While using the RSA framework as the modeling tool requires revising traditional assumptions about the nature of scalar
implicature by relaxing the crisp notion of scalar
implicature as something that is or is not \enquote{calculated} in
interpretation, it provides new flexibility to explicitly discuss
 behavior in experimental settings. One need not adopt the RSA framework as the tool for hypothesizing and testing the link between theoretical constructs and behavior in pragmatic experiments. However, the empirical findings we have reported here---that the inferences researchers draw about ``implicature rate'' are volatile and depend on various features of the paradigm and the linking hypothesis employed--- strongly suggest that experimental pragmatics  as a field must engage more seriously with the foundational questions of what we are measuring in the experiments we run.}

@@

\change{\hypertarget{ref-Yildirim2016}{}
Yildirim, I., Degen, J., Tanenhaus, M.K., \& Jaeger, T.F. (2016). Talker-specificity and adaptation in quantifier interpretation. \emph{Journal of Memory and Language}, \emph{87}, pp. 128-143.}
